---
title: "Multivar stats final project"
author: "Renata Diaz"
date: "11/23/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(vegan)
library(ca)
  
```

## Project plan

1. Get data
    a. load Portal rodent and plant data
    b. summarize by year
    c. standardize according to trapping/survey effort
2. Reduce dimensionality of plant data
    a. try LDA
        1. If base LDA doesn't work, try removing 5% most/least common species
    b. pcoa
    c. separate winter/summer censuses a priori?
3. Distance based redundancy analysis
    a. try to use plant summary axes to predict rodent community
    b. (?) Use LDA rodent topics as ind variable?
4. Variance partitioning to isolate effects of year, plants on rodents

## Get data

### Plant data
* Extracted summer & winter plant censuses for all years 
* Standardized plant abundances according to sampling effort
* Kept seasons separate

### Rodent data
* Rodent censuses on control plots for all years, restricted to granivores
* Standardized according to sampling effort (per census period)
* Summed across all months in each calendar year

## PCoA

* Ran separate principle coordinates analyses on winter and summer datasets.

### Summer PCoA

```{r summer pcoa, echo = T} 
summer_plants <- read.csv('data/summer-plants-adjusted.csv',
                          stringsAsFactors = F)

summer_dist_mat <- vegdist(summer_plants[,2:ncol(summer_plants)], 'bray')
  
summer_pcoa <- cmdscale(summer_dist_mat, k = nrow(summer_plants) - 1, eig = T)
  
# Proportion of variance table 
  eigenvalues <- summer_pcoa$eig[1:nrow(summer_plants)-1]
  propVar <- eigenvalues/sum(eigenvalues)
  cumVar <- cumsum(propVar)
  Summer_PCoA_Table <- cbind(eigenvalues, propVar, cumVar)
  
  Summer_PCoA_Table[1:15,]
  
  
# Scree plot:
  plot(eigenvalues)
  lines(lowess(eigenvalues))
  
```


There seems to be an inflection point in the scree plot around axis 3. 

Moving forward, keeping the first 3 axes as predictor variables for the rodent community. 

```{r save summer axes, echo = F}
summer_vals <- cbind(summer_plants$year, summer_pcoa$points[,1:3])
  
  season_names <- rep("SummerPCoAxis_", 3) %>%
    paste0(1:3)
  season_names <- c('year', season_names)
  colnames(summer_vals) <- season_names
  
  
  write.csv(summer_vals, 'models/summer_pcoa_vals_3.csv', row.names = F)
 
```



### Winter PCoA

```{r winter pcoa, echo = T} 
winter_plants <- read.csv('data/winter-plants-adjusted.csv',
                          stringsAsFactors = F)

winter_dist_mat <- vegdist(winter_plants[,2:ncol(winter_plants)], 'bray')
  
winter_pcoa <- cmdscale(winter_dist_mat, k = nrow(winter_plants) - 1, eig = T)
  
# Proportion of variance table 
  eigenvalues <- winter_pcoa$eig[1:nrow(winter_plants)-1]
  propVar <- eigenvalues/sum(eigenvalues)
  cumVar <- cumsum(propVar)
  winter_PCoA_Table <- cbind(eigenvalues, propVar, cumVar)
  
  winter_PCoA_Table[1:15,]
  
  
# Scree plot:
  plot(eigenvalues)
  lines(lowess(eigenvalues))
  

ordiplot(scores(winter_pcoa)[, c(1, 2)], type = "n", cex = 1, main = "Winter plant PCoA")
## species scores not available
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)

# Add species

species_pc <- wascores(winter_pcoa$points[, 1:2], winter_plants)
text(species_pc, rownames(species_pc), cex = 0.7, col = "red")
```


There seems to be an inflection point in the scree plot around axis 2 or 3. Since stopping at 2 would only capture 39% of variation, going to go for 3. 


```{r save winter axes, echo = F}
winter_vals <- cbind(winter_plants$year, winter_pcoa$points[,1:3])
  
  season_names <- rep("WinterPCoAxis_", 3) %>%
    paste0(1:3)
  season_names <- c('year', season_names)
  colnames(winter_vals) <- season_names
  
  
  write.csv(winter_vals, 'models/winter_pcoa_vals_3.csv', row.names = F)
 
```

Look at axis composition:


```{r winter PC scores}

winter_pc <- wascores(winter_pcoa$points[, 1:3], winter_plants)
winter_pc

  rm(list=ls())

```



## RDA

Redundancy analysis, using combined winter and summer axes and year to predict the rodent community.
 
Restricted to years with both a winter & summer census (n = 27).

``` {r RDA, echo = T}

rodents <- read.csv('data/rodents-adjusted.csv', 
                    stringsAsFactors = F)

summer_axes <- read.csv('models/Summer_pcoa_vals_3.csv', 
                        stringsAsFactors = F)

winter_axes <- read.csv('models/Winter_pcoa_vals_3.csv',
                        stringsAsFactors = F)


pred_vals <- inner_join(winter_axes, summer_axes, by = 'year')

rodents <- filter(rodents, year %in% pred_vals$year) %>%
  select(-year)

rodents_hel <- decostand(rodents, 'hellinger')

rodents_rda <- rda(rodents_hel ~ ., pred_vals)

R2 <- RsquareAdj(rodents_rda)$r.squared
R2adj <- RsquareAdj(rodents_rda)$adj.r.squared

R2
R2adj


anova(rodents_rda, step = 1000)
anova(rodents_rda, by = "axis", step = 1000)


```

Find the most parsimonious model...

```{r reduce vars, echo = T}

set.seed(11)
step.forward <- ordiR2step(rda(rodents_hel ~ 1, data = pred_vals), scope = formula(rodents_rda), 
                           R2scope = F, direction = "forward", pstep = 1000)

# Most parsimonious is Call: rodents_hel ~ Winter 1 + year + Summer 2 + Winter 2 

rod_rda_pars <- rda(rodents_hel ~ WinterPCoAxis_1 + year + SummerPCoAxis_2 + WinterPCoAxis_2 + SummerPCoAxis_1 , pred_vals)

R2p <- RsquareAdj(rod_rda_pars)$r.squared
R2adjp <- RsquareAdj(rod_rda_pars)$adj.r.squared

R2p
R2adjp


anova(rod_rda_pars, step = 1000)
anova(rod_rda_pars, by = "axis", step = 1000)


```

## Variance partitioning

```{r var part, echo = T}
# Variance partitioning
rod_part <- varpart(rodents_hel, ~WinterPCoAxis_1, ~SummerPCoAxis_2, ~WinterPCoAxis_2, ~SummerPCoAxis_1, data = pred_vals)
rod_part

plot(rod_part, digits = 2)
```

WinterPCoAxis_1 has the largest chunk (.4)

### Years - rodents comparison

```{r plot winter pcoa1 v year}

plot(pred_vals$year, pred_vals$WinterPCoAxis_1)
abline(v = 1990, col = 'red')

```