---
title: "Multivar stats final project"
author: "Renata Diaz"
date: "11/23/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(vegan)
library(ca)
  
```


## Get data

* Download rodent and plant data from the Portal Project repo:

```{r download data, eval = F}
source('R/store_rodent_data.R')
source('R/store_plant_data.R')

store_rodent_data(treatment = 'control')
store_plant_data()
```

Adjust data to compensate for irregular trapping and census effort.

### Plant data
* Extracted summer & winter plant censuses for all years 
* Standardized plant abundances according to sampling effort
* Kept seasons separate

### Rodent data
* Rodent censuses on control plots for all years, restricted to granivores
* Standardized according to sampling effort (per census period)
* Summed across all months in each calendar year


```{r adjust data, eval = F}
control_rodents <- read.csv('final-project/data/control-rodents-raw.csv', 
                    stringsAsFactors = F)

summer_plants_c <- read.csv('final-project/data/summer-plants-raw-c.csv', 
                          stringsAsFactors = F)
winter_plants_c <- read.csv('final-project/data/winter-plants-raw-c.csv', 
                          stringsAsFactors = F)
all_plants_c <- rbind(summer_plants_c, winter_plants_c)

store_adjusted_data(plant_data = all_plants_c, rodent_data = control_rodents,
                    focal_type = 'rodent', treatment = 'control')

store_adjusted_data(plant_data = summer_plants_c, focal_type = 'plant',
                    season = 'summer', treatment = 'control')
store_adjusted_data(plant_data = winter_plants_c, focal_type = 'plant',
                    season = 'winter', treatment = 'control')

summer_plants_e <- read.csv('final-project/data/summer-plants-raw-e.csv', 
                            stringsAsFactors = F)
winter_plants_e <- read.csv('final-project/data/winter-plants-raw-e.csv', 
                            stringsAsFactors = F)
all_plants_e <- rbind(summer_plants_e, winter_plants_e)


store_adjusted_data(plant_data = summer_plants_e, focal_type = 'plant',
                    season = 'summer', treatment = 'exclosure')
store_adjusted_data(plant_data = winter_plants_e, focal_type = 'plant',
                    season = 'winter', treatment = 'exclosure')


rm(list=ls())
```


## PCoA

* Transformed raw abundance values using the Wisconsin transformation, and then created a Bray-Curtis dissimilarity matrix for each community.
* Ran PCoA on the dissimilarity matrices, separately for each community. 

### Summer PCoA

```{r summer pcoa, echo = T} 
summer_plants_c <- read.csv('data/summer-control-plants-adjusted.csv',
                          stringsAsFactors = F)

summer_plants_c_wis <- vegan::wisconsin(summer_plants_c[,2:ncol(summer_plants_c)])

summer_dist_mat <- vegdist(summer_plants_c_wis, 'bray')
  
summer_pcoa <- cmdscale(summer_dist_mat, k = nrow(summer_plants_c_wis) - 1, eig = T)


  
# Proportion of variance table 
  eigenvalues <- summer_pcoa$eig[1:nrow(summer_plants_c_wis)-1]
  propVar <- eigenvalues/sum(eigenvalues)
  cumVar <- cumsum(propVar)
  Summer_PCoA_Table <- cbind(eigenvalues, propVar, cumVar)
  
  Summer_PCoA_Table[1:15,]
  
  
# Scree plot:
  plot(eigenvalues)
  lines(lowess(eigenvalues))
  
  

ordiplot(scores(summer_pcoa)[, c(1, 2)], type = "n", cex = 1, main = "Summer plant PCoA")
## species scores not available
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)

# Add species

species_pc <- wascores(summer_pcoa$points[, 1:2], summer_plants_c_wis)
text(species_pc, rownames(species_pc), cex = 0.7, col = "red")
  
```


There seems to be an inflection point in the scree plot around axis 3. 

Moving forward, keeping the first 3 axes as predictor variables for the rodent community. 

```{r save summer axes, echo = F}
summer_vals <- cbind(summer_plants_c$year, summer_pcoa$points[,1:3])
  
  season_names <- rep("SummerPCoAxis_", 3) %>%
    paste0(1:3)
  season_names <- c('year', season_names)
  colnames(summer_vals) <- season_names
  
  
  write.csv(summer_vals, 'models/summer_pcoa_vals_3.csv', row.names = F)
 
```



### Winter PCoA

```{r winter pcoa, echo = T} 
winter_plants_c <- read.csv('data/winter-control-plants-adjusted.csv',
                          stringsAsFactors = F)

winter_plants_c_wis <- vegan::wisconsin(winter_plants_c[,2:ncol(winter_plants_c)])


winter_dist_mat <- vegdist(winter_plants_c_wis, 'bray')
  
winter_pcoa <- cmdscale(winter_dist_mat, k = nrow(winter_plants_c_wis) - 1, eig = T)
  
# Proportion of variance table 
  eigenvalues <- winter_pcoa$eig[1:nrow(winter_plants_c_wis)-1]
  propVar <- eigenvalues/sum(eigenvalues)
  cumVar <- cumsum(propVar)
  winter_PCoA_Table <- cbind(eigenvalues, propVar, cumVar)
  
  winter_PCoA_Table[1:15,]
  
  
# Scree plot:
  plot(eigenvalues)
  lines(lowess(eigenvalues))
  

ordiplot(scores(winter_pcoa)[, c(1, 2)], type = "n", cex = 1, main = "Winter plant PCoA")
## species scores not available
abline(h = 0, lty = 3)
abline(v = 0, lty = 3)

# Add species

species_pc <- wascores(winter_pcoa$points[, 1:2], winter_plants_c_wis)
text(species_pc, rownames(species_pc), cex = 0.7, col = "red")
```


There seems to be an inflection point in the scree plot around axis 2 or 3. Since stopping at 2 would only capture 35% of variation, going to go for 3. 


```{r save winter axes, echo = F}
winter_vals <- cbind(winter_plants_c$year, winter_pcoa$points[,1:3])
  
  season_names <- rep("WinterPCoAxis_", 3) %>%
    paste0(1:3)
  season_names <- c('year', season_names)
  colnames(winter_vals) <- season_names
  
  
  write.csv(winter_vals, 'models/winter_pcoa_vals_3.csv', row.names = F)
 

#rm(list=ls())

```



* Saved the first three axes for both PCoAs.
* Use these six axes, and year, as predictor values for RDA.


## Partial RDA

Partial redundancy analysis, using combined winter and summer axes, conditioned on year, to predict the rodent community.
 
Restricted to years with both a winter & summer census (n = 27).

Used rodent data, summarized yearly, transformed via Hellinger transformation.

``` {r partial RDA, echo = T}

rodents <- read.csv('data/rodents-adjusted.csv', 
                    stringsAsFactors = F)

summer_axes <- read.csv('models/Summer_pcoa_vals_3.csv', 
                        stringsAsFactors = F)

winter_axes <- read.csv('models/Winter_pcoa_vals_3.csv',
                        stringsAsFactors = F)


pred_vals <- inner_join(winter_axes, summer_axes, by = 'year')

rodents <- filter(rodents, year %in% pred_vals$year) %>%
  select(-year)

pred_vals_noy <- select(pred_vals, -year)
pred_vals_y <- select(pred_vals, year)

rodents_hel <- decostand(rodents, 'hellinger')

rodents_prda <- rda(rodents_hel ~ . + Condition(pred_vals_y$year), pred_vals_noy)

R2 <- RsquareAdj(rodents_prda)$r.squared
R2adj <- RsquareAdj(rodents_prda)$adj.r.squared

R2
R2adj


anova(rodents_prda, step = 1000)
anova(rodents_prda, by = "axis", step = 1000)


```

Find the most parsimonious model...

```{r reduce vars part, echo = T}

set.seed(11)
step.forward <- ordiR2step(rda(rodents_hel ~ 1, data = pred_vals), scope = formula(rodents_prda), 
                           R2scope = F, direction = "forward", pstep = 1000)

# Most parsimonious is Call: rodents_hel ~ WinterPCoAxis_1 + WinterPCoAxis_3 + WinterPCoAxis_2 

rod_prda_pars <- rda(rodents_hel ~ WinterPCoAxis_1 + WinterPCoAxis_3 + WinterPCoAxis_2 + Condition(pred_vals_y$year), pred_vals_noy)

pR2p <- RsquareAdj(rod_prda_pars)$r.squared
pR2adjp <- RsquareAdj(rod_prda_pars)$adj.r.squared

pR2p
pR2adjp


anova(rod_prda_pars, step = 1000)
anova(rod_prda_pars, by = "axis", step = 1000)


```


## Regular RDA

```{r regular rda, echo = T}
rodents_rda <- rda(rodents_hel ~ ., pred_vals)

R2 <- RsquareAdj(rodents_rda)$r.squared
R2adj <- RsquareAdj(rodents_rda)$adj.r.squared

R2
R2adj


anova(rodents_rda, step = 1000)
anova(rodents_rda, by = "axis", step = 1000)
```
Find the most parsimonious model...

```{r reduce vars regular, echo = T}

set.seed(11)
step.forward <- ordiR2step(rda(rodents_hel ~ 1, data = pred_vals), scope = formula(rodents_rda), 
                           R2scope = F, direction = "forward", pstep = 1000)

# Most parsimonious is rodents_hel ~ WinterPCoAxis_1 + WinterPCoAxis_3 + WinterPCoAxis_2 +  SummerPCoAxis_2 

rod_rda_pars <- rda(rodents_hel ~ WinterPCoAxis_1 + WinterPCoAxis_3 + WinterPCoAxis_2 + SummerPCoAxis_2, pred_vals)

R2p <- RsquareAdj(rod_rda_pars)$r.squared
R2adjp <- RsquareAdj(rod_rda_pars)$adj.r.squared

R2p
R2adjp


anova(rod_rda_pars, step = 1000)
anova(rod_rda_pars, by = "axis", step = 1000)


```
## Variance partitioning

```{r  partial variance partitioning, echo = T}

rod_ppart <- varpart(rodents_hel, ~ WinterPCoAxis_1, ~ WinterPCoAxis_3, ~  WinterPCoAxis_2, ~ pred_vals_y$year, data = pred_vals_noy)
rod_ppart

plot(rod_ppart, digits = 2)
```

WinterPCoAxis_1 combined with year has the largest chunk (.34); on its own, WinterPCoAxis_1 explains an additional .1

```{r  regular variance partitioning, echo = T}

rod_part <- varpart(rodents_hel, ~ WinterPCoAxis_1, ~ WinterPCoAxis_3, ~  WinterPCoAxis_2, ~ SummerPCoAxis_2, data = pred_vals)
rod_part

plot(rod_part, digits = 2)
```

How strongly is summer axis 2 related to year?

```{r plot summer pcoa2 v year}
plot(pred_vals$year, pred_vals$SummerPCoAxis_2)
```

### Years - rodents comparison

```{r plot winter pcoa1 v year}
 

plot(pred_vals$year, pred_vals$WinterPCoAxis_1)
abline(v = 1990, col = 'red')

colnames(rodents)

rodents_props <- rodents/rowSums(rodents)


plot(pred_vals$year, rodents_props$DS, col = 'blue')
points(pred_vals$year, rodents_props$PE, col = 'green')
points(pred_vals$year, rodents_props$RM, col = 'purple')
points(pred_vals$year, rodents_props$DM, col = 'pink')
abline(v = 1990, col = 'red')




```


I might be totally misinterpreting this, but I'm feeling like this indicates that the 1990 change point (which, if you look at the LDA results, is driven by a decline in RM & PE relative to DM) is a *delayed effect* of DS decline *moderated by winter plant community*???


Why I think this:
* DS changepoint is the mid 80s but DS continue to decline
* Correlated winter plants don't take off until the 90s
* PE/RM too rare to be driving the end of that change
* PE/RM seem to follow the DS into decline
* So if the plant change occurs after the DS change, and PM/RE are tracking the *plants*, the timing seems to make sense. 

Let's look at that same PCOA axis in the *exclosures*...

```{r winter plants exclosures, echo = T}
winter_plants_e <- read.csv('data/winter-exclosure-plants-adjusted.csv',
                          stringsAsFactors = F)
winter_plants_e <- winter_plants_e %>%
  filter(year %in% winter_plants_c$year)

winter_e_zeros <- winter_plants_c

rmcols <- intersect(colnames(winter_plants_e), colnames(winter_e_zeros))
winter_e_zeros <- winter_e_zeros %>%
  filter(year %in% winter_plants_c$year) %>%
  select(-rmcols)
winter_e_zeros <- 0*(winter_e_zeros)

winter_plants_e <- cbind(winter_plants_e, winter_e_zeros)

winter_plants_e_reordered <- winter_plants_e %>%
  select(colnames(winter_plants_c))

winter_plants_e_wis <- vegan::wisconsin(winter_plants_e_reordered[,2:ncol(winter_plants_e_reordered)])



# winter_plants_e_wis <- winter_plants_e_wis %>%
#   select(colnames(winter_plants_c_wis))
# 
# (colnames(winter_plants_e_wis) == colnames(winter_plants_c_wis))

```
